---
title: "featuretablefilter: Feature Table Filtering for Microbiome Data"
author: "Vojtech Barton"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{featuretablefilter: Feature Table Filtering for Microbiome Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

# Introduction

The `featuretablefilter` package provides functions for filtering microbiome feature tables based on coverage and abundance criteria. It is designed to help researchers remove low-quality samples and rare features from their datasets while maintaining comprehensive quality control metrics.

## Key Features

- **Sample Coverage Filtering**: Remove samples with insufficient sequencing depth
- **Feature Abundance Filtering**: Filter out rare features based on absolute or relative thresholds
- **Data-driven Cutoff Estimation**: MAD-based and IQR-based threshold estimation
- **Quality Control Metrics**: Comprehensive metrics including sparsity, retention rates, and rank-abundance stability
- **Visualization**: Multiple plotting functions for exploring filtering effects

# Installation

```{r installation, eval = FALSE}
# Install from local source
devtools::install(".")

# Required packages
library(featuretablefilter)
```

# Loading Data

The package includes a function to load feature tables in common formats:

```{r load-data, eval = FALSE}
# Load TSV file (auto-detects format)
table <- load_feature_table("example_feature_table.tsv")

# Load CSV file
table <- load_feature_table("feature-table.csv")

# The first column should contain feature IDs
head(table)
```

## Feature Table Format

The expected format is a tab-separated or comma-separated file where:
- The first column contains feature IDs (e.g., OTU IDs, ASV IDs, taxon names)
- Remaining columns contain sample counts with sample names as column headers
- Values are integer counts of reads/features per sample

# Sample Coverage Filtering

## Absolute Threshold Filtering

Remove samples with total read count below a fixed threshold:

```{r coverage-absolute, eval = FALSE}
# Filter samples with < 1000 reads
filtered_table <- filter_by_coverage(table, min_reads = 1000)
```

## Data-driven Threshold Estimation

### MAD-based Method

The Median Absolute Deviation (MAD) method identifies outliers using the formula:
$$\text{cutoff} = \max(\text{floor}, \text{median} - \text{multiplier} \times \text{MAD})$$

```{r mad-cutoff, eval = FALSE}
# Estimate cutoff using MAD method (default multiplier = 3)
est <- estimate_mad_cutoff(table, multiplier = 3, floor = 0)
print(est)

# Use the estimated cutoff for filtering
filtered_table <- filter_by_coverage(table, min_reads = est$cutoff)
```

### IQR-based Method

The Interquartile Range (IQR) method uses Tukey's fences:
$$\text{cutoff} = \max(\text{floor}, Q1 - \text{multiplier} \times \text{IQR})$$

```{r iqr-cutoff, eval = FALSE}
# Estimate cutoff using IQR method (default multiplier = 1.5)
est <- estimate_iqr_cutoff(table, multiplier = 1.5, floor = 0)
print(est)

# Use the estimated cutoff for filtering
filtered_table <- filter_by_coverage(table, min_reads = est$cutoff)
```

## Ecological Completeness-Based Filtering

Instead of using absolute sequencing depth cutoffs, you can filter samples based on estimated ecological completeness using coverage estimators. These methods assess whether a sample has been sequenced deeply enough to capture the diversity present.

### Good's Coverage

Good's coverage estimator measures the probability that the next read will belong to a previously observed feature:

$$C = 1 - \frac{n_1}{n}$$

where $n_1$ is the number of singletons (features with exactly one read) and $n$ is the total number of reads.

```{r good-coverage, eval = FALSE}
# Estimate Good's coverage for all samples
good_cov <- estimate_good_coverage(table, target_coverage = 0.95)
print(good_cov)

# Filter to keep only samples with >= 95% estimated coverage
result <- filter_by_coverage_estimator(
  table,
  method = "good",
  target_coverage = 0.95
)
filtered_table <- result$table

# View coverage before and after filtering
cat(sprintf("Mean coverage before: %.2f%%\n", result$mean_coverage_before * 100))
cat(sprintf("Mean coverage after: %.2f%%\n", result$mean_coverage_after * 100))
```

### Chao's Coverage

Chao's coverage estimator is more conservative and accounts for unseen species/features using both singletons and doubletons:

$$\hat{C} = 1 - \frac{f_1}{S} + \frac{f_1}{n} \times \frac{n-1}{n} \times \frac{S-1}{S}$$

where $f_1$ = singletons, $S$ = number of features, $n$ = total reads.

```{r chao-coverage, eval = FALSE}
# Estimate Chao's coverage for all samples
chao_cov <- estimate_chao_coverage(table, target_coverage = 0.90)
print(chao_cov)

# Filter to keep only samples with >= 90% estimated coverage
result <- filter_by_coverage_estimator(
  table,
  method = "chao",
  target_coverage = 0.90
)
filtered_table <- result$table
```

### Comparison of Methods

```{r compare-coverage, eval = FALSE}
# Compare different estimation methods
good_est <- estimate_good_coverage(table, target_coverage = 0.95)
chao_est <- estimate_chao_coverage(table, target_coverage = 0.90)

cat(sprintf("Good's coverage: mean=%.2f%%, min=%.2f%%\n",
            good_est$mean_coverage * 100, good_est$min_coverage * 100))
cat(sprintf("Chao's coverage: mean=%.2f%%, min=%.2f%%\n",
            chao_est$mean_coverage * 100, chao_est$min_coverage * 100))
```

Note: Chao's estimator typically gives lower (more conservative) coverage estimates than Good's estimator.

# Feature Abundance Filtering

## Absolute Count Filtering

Remove features with low total read counts:

```{r feature-absolute, eval = FALSE}
# Filter features with < 5 total reads
filtered_table <- filter_features_by_abundance(
  table,
  threshold = 5,
  mode = "absolute",
  min_samples = 1
)

# Keep only features present in at least 3 samples
filtered_table <- filter_features_by_abundance(
  table,
  threshold = 1,
  mode = "absolute",
  min_samples = 3
)
```

## Relative Abundance Filtering

Filter features based on proportion of total reads:

```{r feature-relative, eval = FALSE}
# Filter features with < 0.1% relative abundance
filtered_table <- filter_features_by_abundance(
  table,
  threshold = 0.001,
  mode = "relative",
  min_samples = 1
)
```

## Joint Abundance-Prevalence Filtering

A powerful filtering approach that combines abundance thresholds with prevalence requirements using AND/OR logic. This is particularly useful for retaining features that are either widely distributed across samples OR have high abundance when present.

### OR Logic (Either Criterion)

Keep features that meet EITHER the abundance threshold OR the prevalence threshold:

```{r joint-or, eval = FALSE}
# Keep features that are EITHER >0.1% relative abundance
# OR present in >=30% of samples
result <- filter_features_joint(
  table,
  abundance_threshold = 0.001,      # 0.1% relative abundance
  prevalence_threshold = 0.3,       # 30% of samples
  mode = "relative",
  logic = "OR"
)

# View detailed breakdown
cat(sprintf("Features before: %d\n", result$n_features_before))
cat(sprintf("Features after: %d\n", result$n_features_after))
cat(sprintf("Met abundance only: %d\n", result$n_by_abundance_only))
cat(sprintf("Met prevalence only: %d\n", result$n_by_prevalence_only))
cat(sprintf("Met both criteria: %d\n", result$n_by_both))
cat(sprintf("Met neither (removed): %d\n", result$n_by_neither))
```

**Use case**: Retain rare but abundant features (e.g., pathogen spikes) AND common but low-abundance features (e.g., core microbiome members).

### AND Logic (Both Criteria)

Keep features that meet BOTH the abundance threshold AND the prevalence threshold:

```{r joint-and, eval = FALSE}
# Keep features that are BOTH >5 reads AND present in >=50% of samples
result <- filter_features_joint(
  table,
  abundance_threshold = 5,
  prevalence_threshold = 0.5,       # 50% of samples
  mode = "absolute",
  logic = "AND"
)
```

**Use case**: Stringent filtering to keep only robust, well-distributed features.

### Using in the Pipeline

```{r pipeline-joint, eval = FALSE}
result <- run_filtering_pipeline(
  input = "example_feature_table.tsv",
  output_dir = "results",
  prefix = "joint_filter",

  # Coverage filtering
  cov_filter_method = "good",
  cov_target_coverage = 0.95,

  # Joint abundance-prevalence filtering
  abun_filter_method = "joint",
  abun_threshold = 0.001,           # 0.1% relative abundance
  abun_prevalence_threshold = 0.3,  # 30% of samples
  abun_logic = "OR",                 # Keep if meets either criterion

  generate_plots = TRUE,
  generate_report = TRUE
)
```

## Relative Cutoff Method

This method calculates an absolute threshold based on a percentage of the minimum-coverage sample:

```{r relative-cutoff, eval = FALSE}
filtered_table <- filter_by_relative_cutoff(
  table,
  min_coverage = 1000,  # Minimum coverage required for a sample
  relative_threshold = 0.01,  # 1% of min-coverage sample
  remove_features = TRUE
)
```

# Complete Filtering Pipeline

The `run_filtering_pipeline()` function orchestrates a complete workflow:

```{r pipeline, eval = FALSE}
result <- run_filtering_pipeline(
  input = "example_feature_table.tsv",
  output_dir = "results",
  prefix = "analysis1",

  # Coverage filtering
  cov_filter_method = "mad",
  cov_threshold = 3,

  # Abundance filtering
  abun_filter_method = "relative_cutoff",
  abun_threshold = 0.01,
  min_coverage_for_relative = 1000,

  # Options
  generate_plots = TRUE,
  generate_report = TRUE,
  verbose = TRUE
)

# Access results
original <- result$original_table
filtered <- result$filtered_table
qc <- result$qc_metrics
```

## Available Methods

### Coverage Filtering Methods

| Method | Description |
|--------|-------------|
| `none` | No coverage filtering |
| `absolute` | Remove samples below fixed read count threshold |
| `mad` | Use MAD-based threshold (median - multiplier * MAD) |
| `iqr` | Use IQR/Tukey-based threshold (Q1 - multiplier * IQR) |
| `good` | Use Good's coverage estimator (ecological completeness) |
| `chao` | Use Chao's coverage estimator (conservative completeness) |

### Abundance Filtering Methods

| Method | Description |
|--------|-------------|
| `none` | No abundance filtering |
| `absolute` | Remove features below fixed count threshold |
| `relative` | Remove features below relative abundance proportion |
| `relative_cutoff` | Relative threshold based on min-coverage sample |
| `joint` | Joint abundance-prevalence filtering with AND/OR logic |

# Quality Control Metrics

Compute comprehensive QC metrics comparing original and filtered tables:

```{r qc-metrics, eval = FALSE}
qc <- compute_filtering_qc(original_table, filtered_table, top_n = 10)

# View results
print(qc)
```

## Metrics Computed

- **Sparsity**: Percentage of zero values in the table
- **Retention Rates**: Lreads, features, and samples retained
- **Rank-Abundance Stability**:
  - Top N feature overlap between original and filtered tables
  - Jaccard similarity index
  - Spearman rank correlation
- **Procrustes Analysis**: Compositional similarity (requires `vegan` package)

# Visualization

## Coverage Distribution

```{r plot-coverage, eval = FALSE}
plot_coverage_histogram(table, threshold = 1000)
```

## Presence-Frequency Analysis

Compare feature prevalence (fraction of samples containing a feature) and sample richness (number of features per sample):

```{r plot-presence, eval = FALSE}
# Single table analysis
plot_presence_analysis(table, threshold = 1)

# Comparison of original vs filtered
plot_presence_analysis(
  original_table,
  table_filtered = filtered_table,
  save_dir = "plots",
  prefix = "presence_comparison"
)
```

## QC Comparison Plots

Generate comprehensive comparison plots:

```{r plot-qc, eval = FALSE}
plot_qc_comparison(
  original_table,
  filtered_table,
  plot_dir = "plots",
  prefix = "qc_comparison",
  top_n = 10,
  heatmap_top = 50
)
```

This generates:
1. Sample coverage distribution (log10 scale, faceted)
2. Feature abundance distribution (log10 scale, faceted)
3. Per-sample sparsity histogram
4. Retention rates barplot
5. Top features stacked barplot
6. Heatmaps of top variable features

## Top Features Stacked Barplot

```{r plot-top-features, eval = FALSE}
plot_top_features_stacked(
  original_table,
  filtered_table,
  top_n = 10,
  plot_dir = "plots",
  prefix = "top_features"
)
```

Note: Top N features are selected independently for each table based on total relative abundance.

# Session Info

```{r session-info}
sessionInfo()
```

# References

For more information, visit the package repository or see the README.md file.
