---
title: "featuretablefilter: Feature Table Filtering for Microbiome Data"
author:
  - name: "Vojtech Barton"
    email: "vojtech.barton@gmail.com"
date: "`r Sys.Date()`"
package: featuretablefilter
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    fig_crop: false
vignette: >
  %\VignetteIndexEntry{featuretablefilter: Feature Table Filtering for Microbiome Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

# Introduction

The `featuretablefilter` package provides functions for filtering microbiome feature tables based on coverage and abundance criteria. It is designed to help researchers remove low-quality samples and rare features from their datasets while maintaining comprehensive quality control metrics.

## Key Features

- **Sample Coverage Filtering**: Remove samples with insufficient sequencing depth
- **Feature Abundance Filtering**: Filter out rare features based on absolute or relative thresholds
- **Data-driven Cutoff Estimation**: MAD-based and IQR-based threshold estimation
- **Quality Control Metrics**: Comprehensive metrics including sparsity, retention rates, and rank-abundance stability
- **Visualization**: Multiple plotting functions for exploring filtering effects

# Installation and Loading

```{r load-package}
library(featuretablefilter)
```

# Loading Data

The package includes a function to load feature tables in common formats. We also ship an example feature table for illustration.

```{r load-data}
# Load example data shipped with the package
data(example_feature_table)
head(example_feature_table[, 1:4])
dim(example_feature_table)

# Load from a TSV file
file_path <- system.file("extdata", "example_feature_table.tsv",
                         package = "featuretablefilter")
table <- load_feature_table(file_path)
head(table[, 1:4])
```

## Feature Table Format

The expected format is a tab-separated or comma-separated file where:

- The first column contains feature IDs (e.g., OTU IDs, ASV IDs, taxon names)
- Remaining columns contain sample counts with sample names as column headers
- Values are integer counts of reads/features per sample

# Sample Coverage Filtering

## Absolute Threshold Filtering

Remove samples with total read count below a fixed threshold:

```{r coverage-absolute}
data(example_feature_table)
filtered_table <- filter_by_coverage(example_feature_table, min_reads = 1000)
ncol(filtered_table)
```

## Data-driven Threshold Estimation

### MAD-based Method

The Median Absolute Deviation (MAD) method identifies outliers using the formula:
$$\\text{cutoff} = \\max(\\text{floor}, \\text{median} - \\text{multiplier} \\times \\text{MAD})$$

```{r mad-cutoff}
data(example_feature_table)
est <- estimate_mad_cutoff(example_feature_table, multiplier = 3, floor = 0)
est$cutoff

filtered_table <- filter_by_coverage(example_feature_table, min_reads = est$cutoff)
ncol(filtered_table)
```

### IQR-based Method

The Interquartile Range (IQR) method uses Tukey's fences:
$$\\text{cutoff} = \\max(\\text{floor}, Q1 - \\text{multiplier} \\times \\text{IQR})$$

```{r iqr-cutoff}
est <- estimate_iqr_cutoff(example_feature_table, multiplier = 1.5, floor = 0)
est$cutoff

filtered_table <- filter_by_coverage(example_feature_table, min_reads = est$cutoff)
ncol(filtered_table)
```

## Ecological Completeness-Based Filtering

Instead of using absolute sequencing depth cutoffs, you can filter samples based on estimated ecological completeness using coverage estimators. These methods assess whether a sample has been sequenced deeply enough to capture the diversity present.

### Good's Coverage

Good's coverage estimator measures the probability that the next read will belong to a previously observed feature:

$$C = 1 - \\frac{n_1}{n}$$

where $n_1$ is the number of singletons (features with exactly one read) and $n$ is the total number of reads.

```{r good-coverage}
good_cov <- estimate_good_coverage(example_feature_table, target_coverage = 0.95)
good_cov$mean_coverage

result <- filter_by_coverage_estimator(
  example_feature_table,
  method = "good",
  target_coverage = 0.95,
  verbose = FALSE
)
ncol(result$table)
```

### Chao's Coverage

Chao's coverage estimator is more conservative and accounts for unseen species/features using both singletons and doubletons:

```{r chao-coverage}
chao_cov <- estimate_chao_coverage(example_feature_table, target_coverage = 0.90)
chao_cov$mean_coverage

result <- filter_by_coverage_estimator(
  example_feature_table,
  method = "chao",
  target_coverage = 0.90,
  verbose = FALSE
)
ncol(result$table)
```

# Feature Abundance Filtering

## Absolute Count Filtering

Remove features with low total read counts:

```{r feature-absolute}
filtered_table <- filter_features_by_abundance(
  example_feature_table,
  threshold = 5,
  mode = "absolute",
  min_samples = 1
)
nrow(filtered_table)
```

## Relative Abundance Filtering

Filter features based on proportion of total reads:

```{r feature-relative}
filtered_table <- filter_features_by_abundance(
  example_feature_table,
  threshold = 0.001,
  mode = "relative",
  min_samples = 1
)
nrow(filtered_table)
```

## Joint Abundance-Prevalence Filtering

A powerful filtering approach that combines abundance thresholds with prevalence requirements using AND/OR logic.

```{r joint-or}
result <- filter_features_joint(
  example_feature_table,
  abundance_threshold = 0.001,
  prevalence_threshold = 0.3,
  mode = "relative",
  logic = "OR"
)
result$n_features_after
```

## Relative Cutoff Method

This method calculates an absolute threshold based on a percentage of the minimum-coverage sample:

```{r relative-cutoff}
filtered_table <- filter_by_relative_cutoff(
  example_feature_table,
  min_coverage = 1000,
  relative_threshold = 0.01,
  remove_features = TRUE
)
nrow(filtered_table)
```

# Complete Filtering Pipeline

The `run_filtering_pipeline()` function orchestrates a complete workflow:

```{r pipeline}
data(example_feature_table)
result <- run_filtering_pipeline(
  input = example_feature_table,
  output_dir = tempdir(),
  prefix = "analysis1",
  cov_filter_method = "mad",
  cov_threshold = 3,
  abun_filter_method = "absolute",
  abun_threshold = 10,
  generate_report = FALSE,
  verbose = FALSE
)

nrow(result$filtered_table)
ncol(result$filtered_table)
```

# Quality Control Metrics

Compute comprehensive QC metrics comparing original and filtered tables:

```{r qc-metrics}
qc <- compute_filtering_qc(example_feature_table, result$filtered_table, top_n = 10)
qc$feature_retention_percent
qc$sample_retention_percent
```

# Visualization

## Coverage Distribution

```{r plot-coverage}
plot_coverage_histogram(example_feature_table)
```

## QC Comparison Plots

```{r plot-qc}
plots <- plot_qc_comparison(example_feature_table, result$filtered_table)
names(plots)
```

## Top Features Stacked Barplot

```{r plot-top-features}
plot_top_features_stacked(example_feature_table, result$filtered_table, top_n = 5)
```

# Session Info

```{r session-info}
sessionInfo()
```

# References

For more information, visit the package repository or see the README.md file.
