---
title: "featuretablefilter Tutorial: Complete Guide to Microbiome Feature Table Filtering"
author:
  - name: "Vojtech Barton"
    email: "vojtech.barton@gmail.com"
date: "`r Sys.Date()`"
package: featuretablefilter
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    fig_crop: false
vignette: >
  %\VignetteIndexEntry{featuretablefilter Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

# Introduction

This tutorial provides a comprehensive guide to using the `featuretablefilter` package for filtering microbiome feature tables. We'll cover:

1. **When and why to filter** - Understanding the need for filtering
2. **Method suitability** - Which methods work best for different sequencing pipelines
3. **Step-by-step pipeline usage** - Complete examples with explanations
4. **Understanding outputs** - How to interpret results and QC metrics
5. **Best practices** - Recommendations based on data type

## Why Filter Feature Tables?

Microbiome sequencing data often contains artifacts that can bias downstream analyses:

- **Low-coverage samples**: Samples with few reads provide unreliable diversity estimates
- **Singleton/doubleton features**: Often PCR or sequencing errors rather than true biological variants
- **Cross-talk/index-hopping**: Reads misassigned between samples on Illumina patterned flow cells
- **Rare features**: May represent contamination or technical artifacts

Proper filtering improves statistical power, computational efficiency, biological interpretability, and reproducibility across studies.

# Loading the Package and Example Data

```{r loading}
library(featuretablefilter)

# Load example feature table shipped with the package
data(example_feature_table)
head(example_feature_table[, 1:4])
dim(example_feature_table)
```

# Loading Your Data

## From File (TSV or CSV)

`load_feature_table()` auto-detects format (TSV vs CSV) and header presence:

```{r load-file}
file_path <- system.file("extdata", "example_feature_table.tsv",
                         package = "featuretablefilter")
table <- load_feature_table(file_path)
head(table[, 1:4])
```

## From phyloseq Object

```{r load-phyloseq, eval = FALSE}
library(phyloseq)
ps <- readRDS("my_phyloseq_object.rds")
table <- from_phyloseq(ps, include_taxa = TRUE)
```

## From TreeSummarizedExperiment

```{r load-tse, eval = FALSE}
library(TreeSummarizedExperiment)
tse <- readRDS("my_tse_object.rds")
result <- from_TSE(tse, add_row_data = TRUE, add_col_data = TRUE)
table <- result$table
```

# Understanding Filtering Methods

## Coverage Filtering Methods

| Method | Best For | Not Recommended For | Description |
|--------|----------|---------------------|-------------|
| `none` | When you want to keep all samples | Low-quality datasets | No coverage filtering |
| `absolute` | Uniform sequencing depth across samples | Highly variable depth | Fixed minimum read count threshold |
| `mad` | Most standard datasets | Very small sample sizes (<10) | Statistical outlier detection using MAD |
| `iqr` | Datasets with non-normal coverage distribution | Small sample sizes | Tukey's fences method |
| `good` | Traditional OTU data, ecological diversity studies | **DADA2/Deblur ASV data** (denoising changes error structure) | Filters by estimated ecological completeness |
| `chao` | Traditional OTU data, conservative rare biosphere studies | **DADA2/Deblur ASV data** (assumes random sampling) | More conservative than Good's coverage |

### Important Notes on Method Selection

#### For DADA2/Deblur Processed Data

If you used **DADA2** or **Deblur** (ASV-level resolution):

- **Avoid aggressive singleton filtering**: ASVs with single reads may be real biological variants, especially in low-biomass samples. DADA2's error modeling already distinguishes true variants from sequencing errors.
- **Use `mad` or `iqr` coverage methods** (NOT Good's/Chao's): Statistical outlier detection is more appropriate because DADA2's denoising changes the error structure and produces variable sequencing depths naturally.
- **Skip cross-talk filtering**: DADA2 already removes chimeric sequences; additional cross-talk filtering may remove real rare variants.

```{r dada2-recommended}
data(example_feature_table)
result <- run_filtering_pipeline(
  input = example_feature_table,
  output_dir = tempdir(),
  prefix = "dada2_example",
  cov_filter_method = "mad",
  cov_threshold = 2,
  abun_filter_method = "joint",
  abun_threshold = 0.0001,
  abun_prevalence_threshold = 0.05,
  abun_logic = "OR",
  crosstalk_filter_method = "none",
  singleton_filter_method = "none",
  generate_report = FALSE,
  verbose = FALSE
)
result$filtering_summary$coverage$n_samples_after
```

#### For OTU Clustering (97% similarity)

OTU tables are more robust to some artifacts but still benefit from filtering:

```{r otu-recommended}
data(example_feature_table)
result <- run_filtering_pipeline(
  input = example_feature_table,
  output_dir = tempdir(),
  prefix = "otu_example",
  cov_filter_method = "mad",
  cov_threshold = 2,
  abun_filter_method = "absolute",
  abun_threshold = 5,
  singleton_filter_method = "absolute",
  singleton_max_ratio = 0.15,
  generate_report = FALSE,
  verbose = FALSE
)
result$filtering_summary$abundance$n_features_after
```

#### For 16S rRNA Amplicon Data (General)

```{r 16s-general}
data(example_feature_table)
result <- run_filtering_pipeline(
  input = example_feature_table,
  output_dir = tempdir(),
  prefix = "16s_example",
  cov_filter_method = "absolute",
  cov_threshold = 1000,
  abun_filter_method = "relative",
  abun_threshold = 0.001,
  generate_plots = TRUE,
  generate_report = FALSE,
  verbose = FALSE
)
result$filtering_summary$coverage$n_samples_after
```

#### For Metagenomic Count Data

Metagenomic feature tables (gene families, pathways) have different characteristics:

```{r metagenomic}
data(example_feature_table)
result <- run_filtering_pipeline(
  input = example_feature_table,
  output_dir = tempdir(),
  prefix = "metagenome_example",
  cov_filter_method = "iqr",
  cov_threshold = 1.5,
  abun_filter_method = "absolute",
  abun_threshold = 10,
  abun_min_samples = 3,
  run_scree_analysis = TRUE,
  generate_report = FALSE,
  verbose = FALSE
)
result$filtering_summary$abundance$n_features_after
```

## Abundance Filtering Methods

| Method | Best For | Description |
|--------|----------|-------------|
| `none` | When keeping all features | No abundance filtering |
| `absolute` | Standard datasets | Remove features below fixed read count |
| `relative` | Normalized/compositional analyses | Remove features below relative abundance |
| `relative_cutoff` | Variable sequencing depth | Threshold based on min-coverage sample |
| `joint` | **Most versatile**, balances sensitivity/specificity | Combined abundance AND/OR prevalence |

# Running the Complete Pipeline

## Basic Usage

```{r basic-pipeline}
data(example_feature_table)
result <- run_filtering_pipeline(
  input = example_feature_table,
  output_dir = tempdir(),
  prefix = "analysis1",
  cov_filter_method = "good",
  cov_target_coverage = 0.95,
  abun_filter_method = "joint",
  abun_threshold = 0.001,
  abun_prevalence_threshold = 0.2,
  abun_logic = "OR",
  singleton_filter_method = "none",
  crosstalk_filter_method = "none",
  sparsity_elbow_detect = TRUE,
  depth_sparsity_detect = TRUE,
  run_scree_analysis = TRUE,
  apply_sparsity_elbow = FALSE,
  apply_depth_sparsity_outliers = FALSE,
  generate_plots = TRUE,
  generate_report = FALSE,
  verbose = FALSE
)

nrow(result$filtered_table)
ncol(result$filtered_table)
```

## Working with phyloseq Objects

```{r phyloseq-pipeline, eval = FALSE}
library(phyloseq)
ps <- readRDS("my_data.rds")
result <- run_filtering_pipeline(
  input = ps,
  prefix = "ps_filtered",
  cov_filter_method = "good",
  abun_filter_method = "joint",
  generate_plots = TRUE,
  generate_report = FALSE,
  verbose = FALSE
)
filtered_ps <- result$filtered_table
class(filtered_ps)
```

## Working with TreeSummarizedExperiment

```{r tse-pipeline, eval = FALSE}
library(TreeSummarizedExperiment)
tse <- readRDS("my_tse.rds")
result <- run_filtering_pipeline(
  input = tse,
  prefix = "tse_filtered",
  cov_filter_method = "mad",
  abun_filter_method = "absolute",
  abun_threshold = 10,
  generate_report = FALSE,
  verbose = FALSE
)
filtered_tse <- result$filtered_table
class(filtered_tse)
```

# Understanding Pipeline Outputs

## Return Value Structure

```{r return-structure}
names(result)
str(result$qc_metrics)
```

## Key Metrics to Report

When publishing, report these metrics:

```{r reporting-metrics}
qc <- result$qc_metrics

cat(sprintf("Initial dataset: %d features, %d samples\\n",
            nrow(result$original_table), ncol(result$original_table) - 1))
cat(sprintf("After filtering: %d features, %d samples\\n",
            nrow(result$filtered_table), ncol(result$filtered_table) - 1))
cat(sprintf("\\nRetention rates:\\n"))
cat(sprintf("  Features: %.1f%% (%d removed)\\n",
            qc$feature_retention_percent,
            nrow(result$original_table) - nrow(result$filtered_table)))
cat(sprintf("  Samples: %.1f%% (%d removed)\\n",
            qc$sample_retention_percent,
            ncol(result$original_table) - ncol(result$filtered_table) - 1))
cat(sprintf("  Reads: %.1f%%\\n", qc$read_retention_percent))
cat(sprintf("\\nSparsity: %.1f%% -> %.1f%% (%.1f pp improvement)\\n",
            qc$sparsity_original * 100,
            qc$sparsity_filtered * 100,
            qc$sparsity_drop_percent))
```

## Using Diagnostic Results

### Sparsity Elbow Detection

Identifies the sequencing depth where ASV discovery crashes:

```{r elbow-usage}
elbow <- result$sparsity_elbow_result
cat(elbow$recommendation)
```

### Depth-Sparsity Outlier Analysis

Finds samples that deviate from expected depth-sparsity relationship:

```{r outlier-usage}
outliers <- result$depth_sparsity_result
cat(sprintf("Detected %d outliers\\n", outliers$n_outliers))
```

# Step-by-Step Manual Filtering

Sometimes you want more control than the pipeline provides.

## Step 1: Explore Your Data

```{r explore-data}
# Coverage per sample
coverages <- colSums(example_feature_table[, -1])
summary(coverages)

# Histogram of coverages
plot_coverage_histogram(example_feature_table, threshold = 1000)

# Calculate sparsity
sparsities <- apply(example_feature_table[, -1], 2, function(x) mean(x == 0))
hist(sparsities, main = "Per-Sample Sparsity", xlab = "Proportion Zeros")
```

## Step 2: Choose Coverage Cutoff

```{r choose-cutoff}
# Option A: MAD-based
mad_est <- estimate_mad_cutoff(example_feature_table, multiplier = 2, floor = 500)
cat(sprintf("MAD recommendation: >= %.0f reads\\n", mad_est$cutoff))

# Option B: Good's coverage
good_est <- estimate_good_coverage(example_feature_table, target_coverage = 0.95)
cat(sprintf("Good's coverage: mean=%.2f%%, min=%.2f%%\\n",
            good_est$mean_coverage * 100, good_est$min_coverage * 100))
```

## Step 3: Apply Coverage Filter

```{r apply-coverage}
filtered <- filter_by_coverage(example_feature_table, min_reads = mad_est$cutoff)
cat(sprintf("Removed %d low-coverage samples\\n",
            attr(filtered, "n_filtered_out")))
```

## Step 4: Choose Abundance Filter

```{r choose-abundance}
scree <- compute_scree(filtered, type = "absolute_feature", n_steps = 10, verbose = FALSE)
plot_scree(scree)
```

## Step 5: Apply Abundance Filter

```{r apply-abundance}
final <- filter_features_by_abundance(
  filtered,
  threshold = 0.001,
  mode = "relative",
  min_samples = 2
)
cat(sprintf("Removed %d rare features\\n",
            attr(final, "n_filtered_out")))
```

## Step 6: Compute QC Metrics

```{r compute-qc}
qc <- compute_filtering_qc(example_feature_table, final)
print(qc[c("sparsity_original", "sparsity_filtered",
           "read_retention_percent", "feature_retention_percent")])
```

# Common Workflows

## Workflow 1: Quick Quality Control

```{r quick-qc}
data(example_feature_table)
result <- run_filtering_pipeline(
  input = example_feature_table,
  output_dir = tempdir(),
  prefix = "quick_qc",
  cov_filter_method = "mad",
  cov_threshold = 2,
  abun_filter_method = "absolute",
  abun_threshold = 5,
  generate_plots = TRUE,
  generate_report = FALSE,
  verbose = FALSE
)
result$qc_metrics$feature_retention_percent
```

## Workflow 2: Conservative Diversity Analysis

```{r conservative}
data(example_feature_table)
result <- run_filtering_pipeline(
  input = example_feature_table,
  output_dir = tempdir(),
  prefix = "conservative",
  cov_filter_method = "good",
  cov_target_coverage = 0.95,
  abun_filter_method = "joint",
  abun_threshold = 0.0001,
  abun_prevalence_threshold = 0.05,
  abun_logic = "OR",
  generate_plots = TRUE,
  generate_report = FALSE,
  verbose = FALSE
)
result$qc_metrics$sample_retention_percent
```

## Workflow 3: Stringent Pathogen Detection

```{r stringent}
data(example_feature_table)
result <- run_filtering_pipeline(
  input = example_feature_table,
  output_dir = tempdir(),
  prefix = "stringent",
  cov_filter_method = "iqr",
  cov_threshold = 2,
  abun_filter_method = "absolute",
  abun_threshold = 50,
  abun_min_samples = 5,
  singleton_filter_method = "absolute",
  singleton_max_ratio = 0.10,
  generate_report = FALSE,
  verbose = FALSE
)
result$qc_metrics$read_retention_percent
```

# Troubleshooting

## All Samples Removed by Coverage Filter

```{r troubleshoot-coverage}
coverages <- colSums(example_feature_table[, -1])
summary(coverages)

result <- filter_by_coverage_estimator(
  example_feature_table,
  method = "good",
  target_coverage = 0.90,
  verbose = FALSE
)
ncol(result$table)
```

## Too Many Features Removed

```{r troubleshoot-features}
result <- filter_features_joint(
  example_feature_table,
  abundance_threshold = 0.001,
  prevalence_threshold = 0.1,
  logic = "OR"
)
result$n_features_after

filtered <- filter_features_by_abundance(
  example_feature_table,
  threshold = 2,
  mode = "absolute",
  min_samples = 2
)
nrow(filtered)
```

## High Sparsity After Filtering

```{r troubleshoot-sparsity}
qc <- compute_filtering_qc(example_feature_table, final)
if (qc$sample_retention_percent < 70) {
  message("Consider relaxing coverage filter")
} else {
  message("Sample retention is acceptable")
}
```

# Best Practices Summary

## Do's

- **Always visualize** your data before filtering (coverage histograms, sparsity plots)
- **Review diagnostic outputs** (scree plots, elbow detection) before applying automatically
- **Document your filtering choices** in methods sections
- **Report retention rates** for features, samples, and reads
- **Use appropriate methods** for your sequencing pipeline (see tables above)
- **Keep raw data** separate from filtered data

## Don'ts

- **Don't apply aggressive singleton filtering** to DADA2/Deblur ASV data
- **Don't filter without understanding** why a method is recommended
- **Don't use default parameters blindly** - adjust based on your data characteristics
- **Don't remove all rare features** - some may be biologically relevant
- **Don't ignore QC metrics** - they tell you if filtering worked as intended

# Session Info

For reproducibility, always record your session info:

```{r session-info}
sessionInfo()
```

# References

- Callahan BJ et al. (2016). DADA2: High-resolution sample inference from Illumina amplicon data. *Nature Methods*.
- McMurdie PJ & Holmes S (2013). phyloseq: An R Package for Reproducible Interactive Analysis and Graphics of Microbiome Census Data. *PLOS ONE*.
- Good TI (1953). The Population Frequencies of Species and the Estimation of Population Parameters. *Biometrika*.

# Additional Resources

- Package documentation: `help(package = "featuretablefilter")`
- Function help: `?run_filtering_pipeline`
- GitHub repository: https://github.com/VojtechBarton/featuretablefilter
- Issue tracker: https://github.com/VojtechBarton/featuretablefilter/issues
