---
title: "featuretablefilter Tutorial: Complete Guide to Microbiome Feature Table Filtering"
author: "Vojtech Barton"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{featuretablefilter Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  eval = FALSE  # Set to TRUE when running with actual data
)
```

# Introduction

This tutorial provides a comprehensive guide to using the `featuretablefilter` package for filtering microbiome feature tables. We'll cover:

1. **When and why to filter** - Understanding the need for filtering
2. **Method suitability** - Which methods work best for different sequencing pipelines
3. **Step-by-step pipeline usage** - Complete examples with explanations
4. **Understanding outputs** - How to interpret results and QC metrics
5. **Best practices** - Recommendations based on data type

## Why Filter Feature Tables?

Microbiome sequencing data often contains artifacts that can bias downstream analyses:

- **Low-coverage samples**: Samples with few reads provide unreliable diversity estimates
- **Singleton/doubleton features**: Often PCR or sequencing errors rather than true biological variants
- **Cross-talk/index-hopping**: Reads misassigned between samples on Illumina patterned flow cells
- **Rare features**: May represent contamination or technical artifacts

Proper filtering improves:
- Statistical power by reducing noise
- Computational efficiency
- Biological interpretability
- Reproducibility across studies

# Installation

```{r installation, eval = FALSE}
# Install from GitHub
devtools::install_github("VojtechBarton/featuretablefilter")

# Load the package
library(featuretablefilter)

# Required dependencies (optional features)
# install.packages(c("phyloseq", "vegan", "igraph"))
# BiocManager::install(c("TreeSummarizedExperiment", "SingleCellExperiment"))
```

# Loading Your Data

## From File (TSV or CSV)

```{r load-file, eval = FALSE}
# Auto-detects format (TSV vs CSV) and header presence
table <- load_feature_table("my_feature_table.tsv")

head(table)
#   FeatureID Sample_1 Sample_2 Sample_3 ...
# 1     ASV_1      125       89      234
# 2     ASV_2       45       67       12
# ...

dim(table)
# [1] 1247   48  # 1247 features, 47 samples (+ 1 feature ID column)
```

## From phyloseq Object

```{r load-phyloseq, eval = FALSE}
library(phyloseq)

# Load your phyloseq object
ps <- readRDS("my_phyloseq_object.rds")

# Convert to feature table format
table <- from_phyloseq(ps, include_taxa = TRUE)

# Preserve the original for later conversion back
original_ps <- ps
```

## From TreeSummarizedExperiment

```{r load-tse, eval = FALSE}
library(TreeSummarizedExperiment)

tse <- readRDS("my_tse_object.rds")

# Extract assay with rowData (taxonomy) if available
result <- from_TSE(tse, add_row_data = TRUE, add_col_data = TRUE)
table <- result$table
sample_metadata <- result$sample_data
```

# Understanding Filtering Methods

## Coverage Filtering Methods

| Method | Best For | Not Recommended For | Description |
|--------|----------|---------------------|-------------|
| `none` | When you want to keep all samples | Low-quality datasets | No coverage filtering |
| `absolute` | Uniform sequencing depth across samples | Highly variable depth | Fixed minimum read count threshold |
| `mad` | Most standard datasets | Very small sample sizes (<10) | Statistical outlier detection using MAD |
| `iqr` | Datasets with non-normal coverage distribution | Small sample sizes | Tukey's fences method |
| `good` | **Diversity studies**, ecological analyses | Presence-absence only studies | Filters by estimated ecological completeness |
| `chao` | **Conservative filtering**, rare biosphere studies | When maximizing sample retention | More conservative than Good's coverage |

### Important Notes on Method Selection

#### For DADA2/Deblur Processed Data

If you used **DADA2** or **Deblur** (ASV-level resolution):

- **Avoid aggressive singleton filtering**: ASVs with single reads may be real biological variants, especially in low-biomass samples
- **Use `good` or `chao` coverage methods**: These preserve samples that have adequate ecological coverage even if total reads are moderate
- **Be cautious with cross-talk filtering**: DADA2 already removes many chimeric sequences; additional cross-talk filtering may remove real rare variants

```{r dada2-recommended, eval = FALSE}
# Recommended for DADA2 data
result <- run_filtering_pipeline(
  input = "dada2_table.tsv",
  cov_filter_method = "good",        # Ecological completeness
  cov_target_coverage = 0.95,        # 95% coverage threshold
  abun_filter_method = "joint",      # Combined abundance-prevalence
  abun_threshold = 0.0001,           # Very low abundance threshold
  abun_prevalence_threshold = 0.1,   # Present in at least 10% of samples
  abun_logic = "OR",                 # Keep if meets either criterion
  crosstalk_filter_method = "none",  # Skip cross-talk for DADA2
  singleton_filter_method = "none"   # Don't filter singletons aggressively
)
```

#### For OTU Clustering (97% similarity)

OTU tables are more robust to some artifacts but still benefit from filtering:

```{r otu-recommended, eval = FALSE}
# Recommended for OTU data
result <- run_filtering_pipeline(
  input = "otu_table.tsv",
  cov_filter_method = "mad",         # MAD-based outlier detection
  cov_threshold = 2,                 # 2 MAD below median
  abun_filter_method = "absolute",   # Remove very rare OTUs
  abun_threshold = 5,                # At least 5 total reads
  singleton_filter_method = "absolute",  # Filter high singleton ratios
  singleton_max_ratio = 0.15         # Max 15% singletons
)
```

#### For 16S rRNA Amplicon Data (General)

```{r 16s-general, eval = FALSE}
# General recommendation for 16S amplicon data
result <- run_filtering_pipeline(
  input = "16s_table.tsv",
  cov_filter_method = "good",
  cov_target_coverage = 0.95,
  abun_filter_method = "relative_cutoff",
  abun_threshold = 0.01,             # 1% of min-coverage sample
  min_coverage_for_relative = 1000,
  generate_plots = TRUE,
  generate_report = TRUE
)
```

#### For Metagenomic Count Data

Metagenomic feature tables (gene families, pathways) have different characteristics:

```{r metagenomic, eval = FALSE}
# For metagenomic count data
result <- run_filtering_pipeline(
  input = "metagenome_counts.tsv",
  cov_filter_method = "iqr",         # IQR handles skewed distributions well
  cov_threshold = 1.5,
  abun_filter_method = "absolute",
  abun_threshold = 10,               # Higher absolute threshold for genes
  abun_min_samples = 3,              # Must appear in at least 3 samples
  run_scree_analysis = TRUE          # Help determine optimal cutoffs
)
```

## Abundance Filtering Methods

| Method | Best For | Description |
|--------|----------|-------------|
| `none` | When keeping all features | No abundance filtering |
| `absolute` | Standard datasets | Remove features below fixed read count |
| `relative` | Normalized/compositional analyses | Remove features below relative abundance |
| `relative_cutoff` | Variable sequencing depth | Threshold based on min-coverage sample |
| `joint` | **Most versatile**, balances sensitivity/specificity | Combined abundance AND/OR prevalence |

# Running the Complete Pipeline

## Basic Usage

```{r basic-pipeline, eval = FALSE}
result <- run_filtering_pipeline(
  # Input specification
  input = "my_feature_table.tsv",    # Or phyloseq/TSE object

  # Output settings
  output_dir = "results/",            # Where to save outputs
  prefix = "analysis1",               # Filename prefix

  # Coverage filtering
  cov_filter_method = "good",         # See table above for options
  cov_target_coverage = 0.95,         # Target 95% ecological coverage

  # Abundance filtering
  abun_filter_method = "joint",
  abun_threshold = 0.001,             # 0.1% relative abundance
  abun_prevalence_threshold = 0.2,    # Present in 20% of samples
  abun_logic = "OR",                  # Keep if meets EITHER criterion

  # Optional filters (default: none)
  singleton_filter_method = "none",
  crosstalk_filter_method = "none",

  # Diagnostic analyses (default: TRUE)
  sparsity_elbow_detect = TRUE,
  depth_sparsity_detect = TRUE,
  run_scree_analysis = TRUE,

  # Apply diagnostic recommendations (default: FALSE - review first!)
  apply_sparsity_elbow = FALSE,
  apply_depth_sparsity_outliers = FALSE,

  # Output options
  generate_plots = TRUE,
  generate_report = TRUE,
  verbose = TRUE
)
```

## Working with phyloseq Objects

```{r phyloseq-pipeline, eval = FALSE}
library(phyloseq)

# Load your phyloseq object
ps <- readRDS("my_data.rds")

# Run pipeline (automatically converts and preserves class)
result <- run_filtering_pipeline(
  input = ps,                          # Pass phyloseq object directly
  prefix = "ps_filtered",
  cov_filter_method = "good",
  abun_filter_method = "joint",
  generate_plots = TRUE,
  generate_report = TRUE
)

# Result filtered_table is ALSO a phyloseq object!
filtered_ps <- result$filtered_table

# Verify class
class(filtered_ps)
# [1] "phyloseq"

# Access components
otu_table(filtered_ps)
tax_table(filtered_ps)
phy_tree(filtered_ps)  # Preserved if present in original
sam_data(filtered_ps)
```

## Working with TreeSummarizedExperiment

```{r tse-pipeline, eval = FALSE}
library(TreeSummarizedExperiment)

tse <- readRDS("my_tse.rds")

result <- run_filtering_pipeline(
  input = tse,
  prefix = "tse_filtered",
  cov_filter_method = "mad",
  abun_filter_method = "absolute",
  abun_threshold = 10
)

# filtered_table is also a TreeSummarizedExperiment
filtered_tse <- result$filtered_table

# Metadata preserved
rowData(filtered_tse)
colData(filtered_tse)
rowTree(filtered_tse)  # If present
```

# Understanding Pipeline Outputs

## Return Value Structure

The pipeline returns a list with these components:

```{r return-structure, eval = FALSE}
str(result)
# List of 9
# $ original_table       : 'phyloseq' num [1:1247, 1:47] ...
# $ filtered_table       : 'phyloseq' num [1:856, 1:42] ...
# $ qc_metrics           :List of 15
#   $ sparsity_original          : num 0.623
#   $ sparsity_filtered          : num 0.412
#   $ sparsity_drop_percent      : num 21.1
#   $ read_retention_percent     : num 94.5
#   $ feature_retention_percent  : num 68.6
#   $ sample_retention_percent   : num 89.4
#   ... (more metrics)
# $ presence_stats       :List of 10
#   $ mean_sample_richness       : num 234
#   $ mean_sample_richness_filt  : num 198
#   ... (more statistics)
# $ filtering_summary    :List of 8
#   $ coverage           :List of 5
#   $ abundance          :List of 7
#   ... (details of each step)
# $ sparsity_elbow_result:List of 7
# $ depth_sparsity_result:List of 8
# $ scree_result         :List of 6
# $ input_class          : chr "phyloseq"
```

## Key Metrics to Report

When publishing, report these metrics:

```{r reporting-metrics, eval = FALSE}
qc <- result$qc_metrics

cat(sprintf("Initial dataset: %d features, %d samples\n",
            nrow(result$original_table), ncol(result$original_table) - 1))
cat(sprintf("After filtering: %d features, %d samples\n",
            nrow(result$filtered_table), ncol(result$filtered_table) - 1))
cat(sprintf("\nRetention rates:\n"))
cat(sprintf("  Features: %.1f%% (%d removed)\n",
            qc$feature_retention_percent,
            nrow(result$original_table) - nrow(result$filtered_table)))
cat(sprintf("  Samples: %.1f%% (%d removed)\n",
            qc$sample_retention_percent,
            ncol(result$original_table) - ncol(result$filtered_table) - 1))
cat(sprintf("  Reads: %.1f%%\n", qc$read_retention_percent))
cat(sprintf("\nSparsity: %.1f%% → %.1f%% (%.1f pp improvement)\n",
            qc$sparsity_original * 100,
            qc$sparsity_filtered * 100,
            qc$sparsity_drop_percent))
```

## Interpreting Generated Files

### Text Report (`prefix_report.txt`)

The pipeline generates a detailed text report:

```
========================================
FEATURE TABLE FILTERING PIPELINE REPORT
========================================

Input file: my_data.tsv
Output directory: results/
Prefix: analysis1
Date: 2026-01-15 14:32:45

--- ORIGINAL TABLE ---
Features: 1247
Samples: 47
Total reads: 2,456,789

--- COVERAGE FILTERING ---
Method: good
Target coverage: 0.95 (95%)
Mean coverage before filtering: 0.9234 (92.34%)
Mean coverage after filtering: 0.9567 (95.67%)
Samples filtered: 5 / 47 (10.6%)

--- ABUNDANCE FILTERING ---
Method: joint
Abundance threshold: 0.001 (0.10%)
Prevalence threshold: 0.20 (20% of samples)
Logic: OR

Feature retention breakdown:
  Met abundance only: 124 features
  Met prevalence only: 89 features
  Met both criteria: 643 features
  Met neither (removed): 391 features

--- QC METRICS ---
Sparsity (original): 62.30%
Sparsity (filtered): 41.20%
Read retention: 94.50%
Feature retention: 68.60%
...
```

### Plots Generated

1. **Coverage distribution** - Log-scaled histogram showing sample depths
2. **Feature abundance distribution** - Distribution of feature frequencies
3. **Sparsity comparison** - Per-sample zero proportions
4. **Retention rates** - Barplot of what was kept/removed
5. **Heatmaps** - Top variable features (before/after)
6. **Top features stacked** - Relative abundance of dominant features
7. **Presence frequency** - Feature prevalence distributions

### Scree Plot

Shows how feature retention changes with different thresholds:

```{r scree-interpretation, eval = FALSE}
# The elbow point indicates where additional filtering
# yields diminishing returns
scree <- result$scree_result

cat(sprintf("Elbow at threshold: %.4f\n", scree$summary$elbow_point$threshold))
cat(sprintf("Features retained at elbow: %.1f%%\n",
            scree$summary$elbow_point$retention_at_elbow))
```

## Using Diagnostic Results

### Sparsity Elbow Detection

Identifies the sequencing depth where ASV discovery crashes:

```{r elbow-usage, eval = FALSE}
elbow <- result$sparsity_elbow_result

cat(elbow$recommendation)
# "Consider filtering out 8 samples (17.0%) below depth 1245 reads..."

# Review before applying!
if (elbow$elbow_threshold > 1000) {  # Only apply if reasonable
  filtered <- filter_by_coverage(original, min_reads = elbow$elbow_threshold)
}
```

### Depth-Sparsity Outlier Analysis

Finds samples that deviate from expected depth-sparsity relationship:

```{r outlier-usage, eval = FALSE}
outliers <- result$depth_sparsity_result

cat(sprintf("Detected %d outliers:\n", outliers$n_outliers))
print(outliers$outliers$sample_name)

# View the plot for manual inspection
# Check plot_<prefix>_depth_sparsity_outliers.png

# Only remove if you confirm they're technical artifacts
```

# Step-by-Step Manual Filtering

Sometimes you want more control than the pipeline provides:

## Step 1: Explore Your Data

```{r explore-data, eval = FALSE}
table <- load_feature_table("my_table.tsv")

# Coverage per sample
coverages <- colSums(table[, -1])
summary(coverages)
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
#    523     890    1245    1389    1678    4567

# Histogram of coverages
plot_coverage_histogram(table, threshold = 1000)

# Calculate sparsity
sparsities <- apply(table[, -1], 2, function(x) mean(x == 0))
hist(sparsities, main = "Per-Sample Sparsity", xlab = "Proportion Zeros")
```

## Step 2: Choose Coverage Cutoff

```{r choose-cutoff, eval = FALSE}
# Option A: MAD-based (statistical outlier detection)
mad_est <- estimate_mad_cutoff(table, multiplier = 2, floor = 500)
cat(sprintf("MAD recommendation: >= %.0f reads\n", mad_est$cutoff))

# Option B: Good's coverage (ecological completeness)
good_est <- estimate_good_coverage(table, target_coverage = 0.95)
cat(sprintf("Good's coverage: mean=%.2f%%, min=%.2f%%\n",
            good_est$mean_coverage * 100, good_est$min_coverage * 100))

# Option C: Visual inspection from histogram
# Look for natural gap in coverage distribution
```

## Step 3: Apply Coverage Filter

```{r apply-coverage, eval = FALSE}
filtered <- filter_by_coverage(table, min_reads = mad_est$cutoff)
cat(sprintf("Removed %d low-coverage samples\n",
            attr(filtered, "n_filtered_out")))
```

## Step 4: Choose Abundance Filter

```{r choose-abundance, eval = FALSE}
# Scree analysis to find optimal threshold
scree <- compute_scree(filtered, type = "mad_multiplier", n_steps = 20)

# View saturation curve
plot_scree(scree)

# Check elbow point
cat(sprintf("Recommended threshold: %.4f\n",
            scree$summary$elbow_point$threshold))
```

## Step 5: Apply Abundance Filter

```{r apply-abundance, eval = FALSE}
final <- filter_features_by_abundance(
  filtered,
  threshold = 0.001,      # 0.1% relative abundance
  mode = "relative",
  min_samples = 2         # Must appear in at least 2 samples
)
cat(sprintf("Removed %d rare features\n",
            attr(final, "n_filtered_out")))
```

## Step 6: Compute QC Metrics

```{r compute-qc, eval = FALSE}
qc <- compute_filtering_qc(table, final)

# Key metrics
print(qc[c("sparsity_original", "sparsity_filtered",
           "read_retention_percent", "feature_retention_percent")])

# Diversity retention (if using Hill numbers)
cat(sprintf("Shannon ENS retention: %.1f%%\n",
            qc$shannon_ens_retention_percent))
```

# Common Workflows

## Workflow 1: Quick Quality Control

For routine quality control with minimal decisions:

```{r quick-qc, eval = FALSE}
result <- run_filtering_pipeline(
  input = "data.tsv",
  cov_filter_method = "mad",
  cov_threshold = 2,
  abun_filter_method = "absolute",
  abun_threshold = 5,
  generate_plots = TRUE,
  generate_report = TRUE
)
```

## Workflow 2: Conservative Diversity Analysis

For studies where retaining rare taxa is important:

```{r conservative, eval = FALSE}
result <- run_filtering_pipeline(
  input = "data.tsv",
  cov_filter_method = "good",
  cov_target_coverage = 0.95,
  abun_filter_method = "joint",
  abun_threshold = 0.0001,
  abun_prevalence_threshold = 0.05,
  abun_logic = "OR",
  generate_plots = TRUE
)
```

## Workflow 3: Stringent Pathogen Detection

For studies focusing on abundant, reliable signals:

```{r stringent, eval = FALSE}
result <- run_filtering_pipeline(
  input = "data.tsv",
  cov_filter_method = "iqr",
  cov_threshold = 2,
  abun_filter_method = "absolute",
  abun_threshold = 50,
  abun_min_samples = 5,
  singleton_filter_method = "absolute",
  singleton_max_ratio = 0.10
)
```

# Troubleshooting

## All Samples Removed by Coverage Filter

```{r troubleshoot-coverage, eval = FALSE}
# Check coverage distribution
coverages <- colSums(table[, -1])
summary(coverages)

# If all samples have low coverage, lower your threshold
# Or use Good's coverage which adapts to your data
result <- filter_by_coverage_estimator(
  table,
  method = "good",
  target_coverage = 0.90  # Lower target
)
```

## Too Many Features Removed

```{r troubleshoot-features, eval = FALSE}
# Use joint filtering with OR logic instead of AND
result <- filter_features_joint(
  table,
  abundance_threshold = 0.001,
  prevalence_threshold = 0.1,
  logic = "OR"  # Less stringent than AND
)

# Or use lower absolute threshold
filtered <- filter_features_by_abundance(
  table,
  threshold = 2,  # Instead of 5 or 10
  mode = "absolute",
  min_samples = 2
)
```

## High Sparsity After Filtering

```{r troubleshoot-sparsity, eval = FALSE}
# This might indicate over-filtering
# Check if you're removing too many samples
qc <- compute_filtering_qc(original, filtered)

if (qc$sample_retention_percent < 70) {
  cat("Consider relaxing coverage filter\n")
}

# Or check for batch effects
# Some batches may have systematically lower coverage
```

# Best Practices Summary

## Do's

✓ **Always visualize** your data before filtering (coverage histograms, sparsity plots)

✓ **Review diagnostic outputs** (scree plots, elbow detection) before applying automatically

✓ **Document your filtering choices** in methods sections

✓ **Report retention rates** for features, samples, and reads

✓ **Use appropriate methods** for your sequencing pipeline (see tables above)

✓ **Keep raw data** separate from filtered data

## Don'ts

✗ **Don't apply aggressive singleton filtering** to DADA2/Deblur ASV data

✗ **Don't filter without understanding** why a method is recommended

✗ **Don't use default parameters blindly** - adjust based on your data characteristics

✗ **Don't remove all rare features** - some may be biologically relevant

✗ **Don't ignore QC metrics** - they tell you if filtering worked as intended

# Session Info

For reproducibility, always record your session info:

```{r session-info}
sessionInfo()
```

# References

- Callahan BJ et al. (2016). DADA2: High-resolution sample inference from Illumina amplicon data. *Nature Methods*.
- McMurdie PJ & Holmes S (2013). phyloseq: An R Package for Reproducible Interactive Analysis and Graphics of Microbiome Census Data. *PLOS ONE*.
- Good TI (1953). The Population Frequencies of Species and the Estimation of Population Parameters. *Biometrika*.

# Additional Resources

- Package documentation: `help(package = "featuretablefilter")`
- Function help: `?run_filtering_pipeline`
- GitHub repository: https://github.com/VojtechBarton/featuretablefilter
- Issue tracker: https://github.com/VojtechBarton/featuretablefilter/issues
